SPP-Perc: Peceptron Learning Based Enhanced Signature Path Prefetcher

### 2018-12-05 related.tex
went through with a chainsaw and hacked off large chunks, then fine-tuned a lot of words. added Elvira's IPC graphs

Perceptron-based prediction mechanisms have proven to yield superior accuracy, 
which represents an opportunity for optimizations in cache management. 
Khan \textit{et. al.} proposed SDBP~\cite{sdbp}, in this work, a sampler 
structure keeps partial tags of sampled sets separate from the cache. 
Three tables of two-bit saturating counters are accessed using a technique 
similar to a skewed branch predictor~\cite{Piece_Linear}. For each hit to 
a block in the sampled set, the program counter (PC) of the relevant memory 
instruction is hashed into the three tables and the corresponding counters 
are decremented. For each eviction from a sampled set, the counters corresponding 
to the PC of the last instruction to access the victim block are incremented. 
For an LLC access, the predictor is consulted by hashing the PC of the memory 
access instruction into the three tables and taking the sum of the indexed 
counters. When the sum exceeds some threshold, the accessed block is 
predicted to be dead. Tags in sampled sets are managed with true LRU and 
a reduced associativity, but the LLC may be managed by any policy. 
The paper applies its predictor to do block replacement and bypass for LLC.

Teran \textit{et al.} propose using perceptron learning for reuse prediction 
driving a bypass and replacement optimization~\cite{Perc_Reuse}. In that work, 
perceptron learning is used to set weights selected by hashes of multiple features 
consisting of the PC, several recent PCs, and two different shifts of the tag 
of the referenced block. Each feature is hashed to index its own table of weights. 
The selected weights are summed and thresholded to make a prediction. When a sampled 
block is reused or evicted, the corresponding weights are decremented or incremented, 
respectively, according to the perceptron learning rule.

The continuation of Perceptron was later introduced in the work, Multiperspective Reuse 
Prediction~\cite{Multiperspective}. Compared to previous work that focuses on one or two 
features at a time, Multiperspective prediction uses many features, each contributing 
to the overall prediction. The prediction enables placement, promotion and bypass 
optimizations that further improves performance.

@@

Two widely used prefetchers are Next-n Lines~\cite{nextn} and Stride~\cite{stride}
prefetchers, both of which capture regular memory access patterns with very low hardware overhead. 
The Next-n Lines prefetcher simply queues prefetches for the next n lines after any given miss, 
under the expectation that the principle of spatial locality will hold and those cache lines after 
a missed line are likely to be used in the future. The stride prefetcher is slightly more 
sophisticated -- it attempts to identify simple stride reference patterns in programs based 
upon the past behavior of missing loads. Similar to the Next-n Lines technique, 
when a given load misses, cache lines ahead of that miss are fetched in the pattern 
following the previous behavior in the hope of avoiding future misses. Both prefetchers 
have been widely used due to the simple design and low hardware overhead. However, 
without further knowledge about temporal locality and application characteristics, 
these prefetchers cannot do more than detecting and prefetching regular memory access 
patterns with limited spatial locality.

Somogyi et al. proposed one of the top practical, low-overhead prefetchers, the Spatial 
Memory Streaming (SMS) prefetcher~\cite{SMS}. SMS leverages code-based correlation 
to take advantage of spatial locality in the applications over larger regions of memory 
(called spatial regions). It predicts the future access pattern within a spatial 
region around a miss, based on a history of access patterns initiated by that 
missing instruction in the past. While the SMS prefetcher is effective, it is 
indirectly inferring future program control-flow when it speculates on the misses 
in a spatial region. As a result, the state overheads of this predictor can 
be higher than the others in this class. This idea was further extended by 
the class of related prefetchers~\cite{STMS,Temporal_Instruction_Fetch}.

While previous work had proposed dynamically finding the best prefetch offset, they have failed to 
look into prefetch timeliness. Michaud proposed the Best-Offset prefetcher~\cite{BOP}. 
A prefetch offset is set automatically and dynamically tries to adapt to the application behavior. 
A good prefetch offset is determined by looking at two recent access to a particularly line 
that did not happen too recently. The time between both accesses should be 
greater than the latency for completing a prefetch request.

\subsection{Lookahead Prefetchers}
While the stride and the next-n perefetchers are efficient, they suffer from their limited
ability to capture memory delta patterns. Moreover, a good prefetch proves to be useful only 
if the prefetcher manages to bring the block before its next demand access. Both these issues 
can be overcome by Lookahead Prefetchers. These prefetchers try to use their own predictions 
to create the spelative path of the program's memory access pattern. By going down the speculation
trail till a certain depth, the prefetcher can bring in cache lines well before they are demand 
accessed by the processor. VLDP~\cite{VLDP} prefetches a static depth ahead of the demand fetch, 
without considering the prefetching confidence. More recent Lookahead Prefetchers include SPP and
KPC~\cite{KPC}

Talk about DA-AMPM

@@

A low-accuracy aggressive prefetcher can significantly harm performance. 
In order to minimize the interference from prefetching, Wu \textit{et. al} proposed PACMan~\cite{pacman}, 
a prefetch-aware cache management policy. PACMan dedicated a few sets of the LLC to each of the three 
competing policies that treat demand and prefetch request differently and uses the policy that 
shows the lowest number of cache misses. Similarly, Seshadri  \textit{et. al.} proposed ICP~\cite{icp}. 
ICP is also designed as a comprehensive mechanism to mitigate prefetching interference. It simply 
demotes a prefetch to the lowest priority on a demand hit based on the observation that most 
prefetches are dead after their first hit. To address prefetcher-caused cache pollution, it also 
uses a variation of EAF~\cite{eaf} to track prefetching accuracy and inserts only an accurate 
prefetch to the higher priority position in the LRU stack. ICP assumes, however, that all 
prefetches are inserted only into the LLC, which restricts the maximum benefit of prefetching.

\subsection{Machine Learning for Prefetching}
Peled \textit{et.al.} introduced
interesting ideas for on-line Reinforcement Learning and dynamically scaling
the magnitude of feedback given to the baseline prefetcher~\cite{Semantics}.
The biggest challenge here is that the prefetcher relies on compiler support for
getting features to build the context.

Liao \textit{et. al.} focused on prefetching for data center
applications~\cite{Datacenter}. They used offline machine learning algorithms
such as SVMs and logistic regression to do a parametric search for an optimal
prefetcher configuration.

>>>>

Perceptron-based prediction mechanisms have proven to yield superior accuracy, 
which represents an opportunity for optimizations in cache management. 
%Khan \textit{et. al.} proposed SDBP~\cite{sdbp}, in this work, a sampler 
%structure keeps partial tags of sampled sets separate from the cache. 
%Three tables of two-bit saturating counters are accessed using a technique 
%similar to a skewed branch predictor~\cite{Piece_Linear}. For each hit to 
%a block in the sampled set, the program counter (PC) of the relevant memory 
%instruction is hashed into the three tables and the corresponding counters 
%are decremented. For each eviction from a sampled set, the counters corresponding 
%to the PC of the last instruction to access the victim block are incremented. 
%For an LLC access, the predictor is consulted by hashing the PC of the memory 
%access instruction into the three tables and taking the sum of the indexed 
%counters. When the sum exceeds some threshold, the accessed block is 
%predicted to be dead. Tags in sampled sets are managed with true LRU and 
%a reduced associativity, but the LLC may be managed by any policy. 
%The paper applies its predictor to do block replacement and bypass for LLC.

Teran \textit{et al.} propose perceptron learning for reuse prediction for
bypass and replacement~\cite{Perc_Reuse}. Perceptron learning trains weights
selected by hashes of multiple features consisting of the PC of the memory
access instruction, some recent PCs, and two different shifts of the tag of
the referenced block. Features are hashed to index weights tables, reading out
weights that are and thresholded to make a prediction. When a block from one
of a few sampled sets~\cite{sdbp} is reused or evicted, corresponding weights
are decremented or incremented, respectively according to the perceptron
learning rule.  Multiperspective Reuse Prediction~\cite{Multiperspective}
improves on Perceptron by contributing many new features.

%features. Compared to previous work that focuses on one or two features at a
%time, Multiperspective prediction uses many features, each contributing to the
%overall prediction. The prediction enables placement, promotion and bypass
%optimizations that further improves performance.

@@

Two widely used prefetchers are Next-$n$ Lines~\cite{nextn} and
Stride~\cite{stride} prefetchers. Both capture regular memory access patterns
with low overhead. The Next-$n$ Lines prefetcher queues prefetches for the
next $n$ blocks after any given miss, expecting that the principle of spatial
locality will hold and those cache blocks after a missed block are likely to
be used in the future. The stride prefetcher identifies strided reference
patterns in programs based on past behavior of missing loads.  When a load
misses, cache blocks ahead of that miss are fetched in the pattern following
the previous behavior in the hope of avoiding future misses.  Without further
knowledge about temporal locality and application characteristics, these
prefetchers cannot do more than detecting and prefetching regular memory
access patterns with limited spatial locality.

Somogyi et al. proposed a practical, low-overhead prefetcher: Spatial Memory
Streaming (SMS) prefetcher~\cite{SMS}. SMS leverages code-based correlation to
take advantage of spatial locality in the applications over larger regions of
memory called spatial regions. It predicts future access pattern within a
spatial region around a miss based on a history of access patterns initiated
by that missing instruction in the past. SMS is effective, but it indirectly
infers future program control-flow when it speculates on the misses in a
spatial region. Thus, the state overhead of this predictor can be higher than
the others in this class. The idea was further extended by the class of
related prefetchers~\cite{STMS,Temporal_Instruction_Fetch}.

Other work also considers prefetch timeliness. Michaud proposed the
Best-Offset prefetcher~\cite{BOP}.  A prefetch offset is set automatically and
dynamically tries to adapt to the application behavior.  A good prefetch
offset is determined by looking at two recent access to a particular block
that did not happen too recently. The time between both accesses should be
greater than the latency for completing a prefetch request.

\subsection{Lookahead Prefetchers}
Previous prefetchers have only limited ability to capture memory delta
patterns. Moreover, a good prefetch proves to be useful only if the prefetcher
manages to bring the block before its next demand access. Both these issues
are overcome by Lookahead Prefetchers that use their own predictions to create
the speculative path of the program's memory access pattern. By going down the
speculation trail to a certain depth, the prefetcher can bring in cache blocks
well before they are demand accessed by the processor. VLDP~\cite{VLDP}
prefetches a static depth ahead of the demand fetch, without considering the
prefetching confidence. More recent Lookahead Prefetchers include SPP and
KPC~\cite{KPC}

%Talk about DA-AMPM

@@

A low-accuracy aggressive prefetcher can significantly harm performance.  To
minimize interference from prefetching, Wu \textit{et. al} proposed
PACMan~\cite{pacman}, a prefetch-aware cache management policy.  PACMan
dedicates some LLC sets to each of three competing policies that treat demand
and prefetch requests differently, using the policy that shows the fewest
misses. Seshadri \textit{et. al.} proposed ICP~\cite{icp} that is also
designed as a comprehensive mechanism to mitigate prefetching interference. It
demotes a prefetch to the lowest priority on a demand hit based on the
observation that most prefetches are dead after their first hit.  To address
prefetcher-caused cache pollution, it also uses a variation of EAF~\cite{eaf}
to track prefetching accuracy and inserts only an accurate prefetch to the
higher priority position in the LRU stack.

\subsection{Machine Learning for Prefetching}
Peled \textit{et.al.} introduced interesting ideas for on-line Reinforcement
Learning and dynamically scaling the magnitude of feedback given to the
baseline prefetcher~\cite{Semantics}.  The prefetcher relies on compiler
support for getting features to build the context.  Liao \textit{et. al.}
focused on prefetching for data center applications~\cite{Datacenter}. They
used offline machine learning algorithms such as SVMs and logistic regression
to do a parametric search for an optimal prefetcher configuration.

### 2018-12-05 results.tex
went through with a chainsaw and hacked off large chunks, then fine-tuned a lot of words. added Elvira's IPC graphs

\begin{figure*}[ht]
\includegraphics[width=\textwidth]{4Core_SPEC2017}
\caption{Normalized Speedup for 4-core SPEC CPU 2017 Workloads}
\label{Fig:4Core_SPEC2017}
\end{figure*}

@@

\noindent \textbf{4-CORE ENVIRONMENT}
\newline
Figure~\ref{Fig:4Core_SPEC2017}(a) shows a
comparison of speedups obtained on 4-core mixes of a memory intensive subset
SPEC CPU 2017.  We plot all 4 prefetchers, normalized to the baseline.  The
workloads have been sorted in increasing order of the speedup.  PPF offers a
speedup of \textbf{51.7\%} on these traces, and improvement of \textbf{11.9\%}
over the baseline SPP, \textbf{10.2\%} over the next DA-AMPM, and
\textbf{17.4\%} over BOP.

On a different set of fully random SPEC CPU 2017 4-core mixes, PPF provides an
IPC speedup of \textbf{26.2}\% over the baseline, which is an improvement of
\textbf{5.7}\% over SPP.
\newline

\begin{figure*}[ht]
\includegraphics[width=\textwidth]{8Core_SPEC2017}
\caption{Normalized Speedup for 8-core SPEC CPU 2017 Workloads}
\label{Fig:8Core_SPEC2017}
\end{figure*}

\noindent \textbf{8-CORE ENVIRONMENT}
\newline
The sorted comparison of speedups on the memory
intensive 8-core mixes is shown in Figure~\ref{Fig:8Core_SPEC2017}.  PPF
improves baseline performance by \textbf{38.6\%}, an improvement of
\textbf{10.7\%} over SPP.  For a random set of SPEC CPU 2017 mixes, PPF
improves performance by \textbf{23.6\%} over the baseline, corresponding to
\textbf{4.8\%} over SPP.  This increased improvement achieved by PPF over the
base engine SPP in a multi-core environment is expected as PPF has a very
accurate filter, it eliminates useless prefetches before they can cause
pollution in the shared LLC.

@@

\begin{figure}[ht]
\begin{adjustwidth}{-1cm}{}
  \includegraphics[width=1.2\columnwidth]{AddnConstr}
  \caption{IPC speedup for Small LLC and Low BW}
  \label{Fig:AddnConstr}
\end{adjustwidth}
\end{figure}

@@

Figure~\ref{Fig:AddnConstr}(a) and (b) show the performance of PPF in a
reduced LLC and with low bandwidth constraints, respectively. For the sake of
brevity, we have only shown selected traces followed by the overall
performance on the complete benchmark.  Benchmark {\tt 605.mcf\_s} in low
bandwidth conditions is prefetch averse.  In general, any prefetcher yields a
negative speedup on that trace.  On {\tt 654.roms\_s} and {\tt
607.cactuBSSN\_s}, PPF is unable to match the performance achieved by the best
prefetcher. On the other hand, PPF outperforms all the other prefetchers on
{\tt 623.xalancbmk\_s} and {\tt 638.imagick\_s} benchmarks.  Overall, PPF gives
a better improvement under small LLC condition and matches the best
prefetcher, BOP, under low DRAM bandwidth conditions.

>>>>

%\begin{figure*}[ht]
%\includegraphics[width=\textwidth]{4Core_SPEC2017}
%\caption{Normalized Speedup for 4-core SPEC CPU 2017 Workloads}
%\label{Fig:4Core_SPEC2017}
%\end{figure*}

\begin{figure}[ht]
\includegraphics[width=2.75in]{4CORE_IPC_PPF}
\caption{Speedup for 4-core SPEC CPU 2017}
\label{Fig:4Core_SPEC2017}
\end{figure}

@@

\noindent \textbf{4-CORE ENVIRONMENT}
\newline
Figure~\ref{Fig:4Core_SPEC2017} shows a comparison of speedups obtained on
4-core mixes of a memory intensive subset SPEC CPU 2017.  We plot all 4
prefetchers, normalized to the baseline.  The workloads have been sorted in
increasing order of the speedup.  PPF offers a speedup of \textbf{51.7\%} on
these traces, and improvement of \textbf{11.9\%} over the baseline SPP,
\textbf{10.2\%} over the next DA-AMPM, and \textbf{17.4\%} over BOP.

On a different set of fully random SPEC CPU 2017 4-core mixes (not illustrated
for space reasons), PPF provides an IPC speedup of \textbf{26.2}\% over the
baseline, which is an improvement of \textbf{5.7}\% over SPP.

%\begin{figure*}[ht]
%\includegraphics[width=\textwidth]{8Core_SPEC2017}
%\caption{Normalized Speedup for 8-core SPEC CPU 2017 Workloads}
%\label{Fig:8Core_SPEC2017}
%\end{figure*}

\begin{figure}[ht]
\includegraphics[width=2.75in]{8CORE_IPC_PPF}
\caption{Speedup for 8-core SPEC CPU 2017}
\label{Fig:8Core_SPEC2017}
\end{figure}

\noindent \textbf{8-CORE ENVIRONMENT}
\newline
The sorted comparison of speedups on the memory intensive 8-core mixes is
shown in Figure~\ref{Fig:8Core_SPEC2017}.  PPF improves baseline performance
by \textbf{38.6\%}, an improvement of \textbf{10.7\%} over SPP.  For a random
set of SPEC CPU 2017 mixes (not illustrated for space reasons), PPF improves
performance by \textbf{23.6\%} over the baseline, corresponding to
\textbf{4.8\%} over SPP.  This increased improvement achieved by PPF over the
base engine SPP in a multi-core environment is expected as PPF has a very
accurate filter, it eliminates useless prefetches before they can cause
pollution in the shared LLC.

@@

% djimenez: cut this graph and modified text accordingly
%\begin{figure}[ht]
%\begin{adjustwidth}{-1cm}{}
%  \includegraphics[width=1.2\columnwidth]{AddnConstr}
%  \caption{IPC speedup for Small LLC and Low BW}
%  \label{Fig:AddnConstr}
%\end{adjustwidth}
%\end{figure}

@@

%Figure~\ref{Fig:AddnConstr}(a) and (b) show the performance of PPF in a
%followed by the overall performance on the complete benchmark.  
%For the sake of brevity, we have only shown selected traces

We also model PPF with reduced LLC and with low bandwidth constraints,
respectively (not illustrated for space reasons).  Benchmark {\tt 605.mcf\_s}
in low bandwidth conditions is prefetch averse.  In general, any prefetcher
yields a negative speedup on that trace.  On {\tt 654.roms\_s} and {\tt
607.cactuBSSN\_s}, PPF is unable to match the performance achieved by the best
prefetcher. On the other hand, PPF outperforms all the other prefetchers on
{\tt 623.xalancbmk\_s} and {\tt 638.imagick\_s} benchmarks.  Overall, PPF
gives a better improvement under small LLC condition and matches the best
prefetcher, BOP, under low DRAM bandwidth conditions.
